### 2021-05-14 submission/manuscript.tex
Add abstract, methods, discussion

(Modify depending on target journal, currently abstract submitted to
World Microbe Forum) Antibiotics are a major risk factor for
\emph{Clostridioides difficile} infections (CDIs) because of their
impact on the intestinal microbiome. However, non-antibiotic medications
such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350,
also alter the microbiota, but whether PEG impacts CDI susceptibility
and clearance is unclear. To examine how PEG impacts susceptibility, we
treated C57Bl/6 mice with 5-day and 1-day doses of 15\% PEG in the
drinking water and then challenged the mice with \emph{C. difficile} 630
spores. We used clindamycin-treated mice as a control because they
consistently clear \emph{C. difficile} within 10 days post-challenge
(dpc). To examine how PEG treatment impacts clearance, we administered
PEG for 1 day to clindamycin-treated, \emph{C. difficile}-challenged
mice either immediately following challenge or 3 dpc. We collected
longitudinal stool samples to examine \emph{C. difficile} levels in the
stool via anaerobic culture and profiled the microbiota by 16S rRNA
sequencing. PEG treatment alone was sufficient to render mice
susceptible to CDI and 5-day PEG-treated mice remain colonized for up to
30 dpc. Additionally, 5-day PEG treated mice remained susceptible to CDI
10-days post treatment. In contrast, 1-day PEG treated mice were
transiently colonized, clearing \emph{C. difficile} within 7 dpc.
Although 5-day PEG-treated mice exhibited prolonged \emph{C. difficile}
colonization, we saw no difference in histological inflammation between
PEG- and clindamycin-treated mice. Additionally, administering PEG to
mice after \emph{C. difficile} challenge prolonged colonization up to 30
dpc in mice that received PEG immediately after challenge and 15 dpc in
mice that received PEG 3 dpc. When we examined microbiota composition
across our different treatment groups, we found decreased richness in
the PEG-treated mice that exhibited prolonged \emph{C. difficile}
colonization. Importantly, there were increased Bacteroides and
Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in
most of the PEG-treated mice with prolonged \emph{C. difficile}
colonization. Our findings suggest the osmotic laxative PEG 3350 alters
the mouse microbiota and disrupts colonization resistance to \emph{C.
difficile}, as well as clearance in mice with a CDI. Considering that
most hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact human
microbiota colonization resistance.

@@

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because they disrupt microbiota colonization
resistance (1). However, antibiotics are not the only types of
medications that disrupt the microbiota (2--4). Although, other
medications (proton pump inhibitors, osmotic laxatives, antimotility
agents, and opioids) have been implicated as risk or protective factors
for CDIs through epidemiological studies, whether the association is due
to their impact on the microbiome is still unclear (5--9).

@@

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} infection (14--17).
The mucus barrier is thought to mediate protection from \emph{C.
difficile} infections by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). Whether laxative
administration results in more severe CDIs in mice and how long mice
remain colonized with \emph{C. difficile} after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although, two groups have now shown PEG treatment
alone renders mice susceptible to \emph{C. difficile} (15, 17), these
studies have raised additional questions regarding the dynamics and
severity of infection as well as the role of laxative treatment in
\emph{C. difficile} clearance. Addressing these questions will better
inform how we think about laxatives in the context of CDIs. Here, we
characterized how long PEG-treated mice remain susceptible, whether PEG
treatment results in more sustained \emph{C. difficile} colonization and
severe CDI and than mice treated with clindamycin, and if PEG treatment
after colonization can promote \emph{C. difficile} clearance.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders the mice susceptible to \emph{C. difficile} colonization, with
mice typically clearing \emph{C. difficile} within 10 days
post-challenge (9, 26). All three groups of PEG-treated mice were
administered a 15\% PEG solution in the drinking water for 5-days, 1.
received no additional treatment, 2. was also treated with clindamycin,
and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A).
PEG treatment resulted in weight loss in all 3 groups of PEG-treated
mice relative to their baseline weights, with the greatest change in
weight observed on the fifth day of PEG treatment. The mice recovered
most of the lost weight by five days after treatment (Fig. 1B). After
either the PEG, clindamycin, or PEG and clindamycin treatment all mice
were challenged with 10\textsuperscript{5} \emph{C. difficile} 630
spores (Fig. 1A). All treatments rendered mice susceptible to \emph{C.
difficile} colonization. However, PEG-treated mice remained colonized
with \emph{C. difficile} at a high level through thirty days
post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice
cleared \emph{C. difficile} within ten days post-challenge (Fig. 1C).
Therefore, PEG treatment led to sustained colonization in contrast to
clindamycin mice that naturally cleared \emph{C. difficile} within ten
days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to
\emph{C. difficile} colonization after a 10-day recovery period,
although \emph{C. difficile} was not detectable in most of the group in
the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
dead on the 6th day post-challenge, presumable due to \emph{C.
difficile} as the bacteria became detecable in that mouse on the 4th day
post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward,
the density of \emph{C. difficile} stabilized in the 10-day recovery
group and remained high through 20-30 days post-challenge (Fig. 1C).
Thus, osmotic laxative treatment alone was sufficient to render mice
susceptible to prolonged \emph{C. difficile} colonization and
PEG-treated mice remained susceptible through ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupted the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation plate
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to coprophagy (27). We tried to
minimize the impact of cage effects on our experiment by breaking up
cagemates when assigning mice to treatment groups. Importantly, although
we conducted a total of 5 separate experiments, the experiment number
and its interaction with treatment group was not one of the variables
that significantly explained the observed variation in fecal communities
(Data Set S1, Sheet X). None of the treatment groups recovered to their
baseline community structure either 10 or 30 days post-challenge
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

@@

Interestingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine if there were genera that
changed during the post-challenge period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-challenge to
detectable in the stool samples with the density stabilizing around 8
days post-challenge (Fig. S1A). We found no bacteria were significantly
different over the two time periods after multiple hypothesis correction
(Data Set S1, Sheet X). However, there was also wide variation between
individual mice regarding when \emph{C. difficile} became detectable
(Fig. S1A) as well as the relative abundances of bacterial genera in the
communities (Fig. S1B). For example, two mice had a high relative
abundance of \emph{Enterobacteriaceae} throughout the post-challenge
period and this corresponded to mouse 10, which died on the sixth day
post-challenge and mouse 11, where \emph{C. difficile} was present at a
high density from the 4th day post-challenge onward (Fig. S1B). While we
did not identify a clear signal to explain the delayed appearance of
\emph{C. difficile} in the 5-day PEG mice that were allowed to recover
for 10 days prior to challenge, the delay is striking and could reflect
changes in microbial activity or metabolites that were not examined in
this study.

\textbf{5-day laxative treatment did not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced snips of tissue collected
from the cecum, proximal colon, and distal colon. Similar to what was
observed with the stool samples, alpha diversity was lower in the
PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
diversity continued to increase over time with the PEG-treated mice
collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet
X). Group, time point, and their interactions with other variables
(cage, experiment number, and sample type) explained the majority of the
variation observed in mucosal communities (PERMANOVA combined
R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
Set S1, Sheet X). We saw the greatest difference in the relative
abundance of the mucosal microbiota between treatment groups
(clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
post-challenge with 10 genera that were significantly different
(\emph{P} `r format.pval(tissue\_genera\_max\_p, eps = 0.05)) in all
three of the tissue types we collected (cecum, proximal colon, and
distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly,
\emph{Peptostreptococcaceae} (the family with a sequence that matches
\emph{C. difficile}) was one of the genera that had a significant
difference in relative abundance between treatment groups at 6 days
post-challenge. This population was primarily only present in the 5-day
PEG treatment group of mice and decreased in the proximal and distal
colon tissues over time (Fig. S2B). By 30 days post-challenge, only the
relative abundances of \emph{Bacteroides}, \emph{Clostridiales},
\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data
Set S1, Sheet X). Thus, PEG treatment had a significant impact on the
mucosal microbiota and we detected \emph{C. difficile} sequences in the
cecum, proximal colon, and distal colon tissue communities.

Next, we examined the severity of \emph{C. difficile} challenge by
evaluating cecum and colon H\&E stained histopathology (28) and found
there was no difference in cecum and colon scores between clindamycin
and PEG-treated mice that were challenged with \emph{C. difficile} at 4
days post-challenge (Fig. 3D), the time point typically examined in
\emph{C. difficile} 630 challenged mice (29). We also looked at 6 days
post-challenge because that was when there was a large difference in
\emph{C. difficile} density between PEG- and clindamycin-treated mice
(Fig. 1C). Although, there was a slight difference in the colon between
PEG and clindamycin-treated mice, there was not a signifant difference
in the cecum and the overall score was relatively low (1.5-2.5 out of
12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect
on the mucosal microbiota, the impact of \emph{C. difficile} 630
challenge on the cecum and colon was similar between PEG and clindamycin
treated mice.

@@

\textbf{Post-CDI laxative treatment disrupted clearance in
clindamycin-treated mice regardless of whether an FMT was also
administered.} Since a 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. To examine the impact of PEG treatment on
\emph{C. difficile} clearance, we treated 4 groups of mice with
clindamycin and then challenged all mice with \emph{C. difficile} before
administering the following treatments: no additional treatment, 1-day
PEG immediately after challenge, and 1-day PEG treatment 3 days after
challenge followed by either administration of an FMT or PBS solution by
oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice
that received PEG exhibited prolonged \emph{C. difficile} colonization
(Fig. 5B). Next we examined how post-CDI PEG treatment impacted
microbiota diversity. Alpha diversity was lower in the PEG-treated mice
with the exception of the PEG-treated mice that were administered an FMT
(Fig 5C-D).

@@

\textbf{Five-day post-challenge community data can predict which mice
that will have prolonged \emph{C. difficile} colonization.} After
identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day
PEG treatments, we decided to examine the bacteria that influenced
prolonged \emph{C. difficile} colonization. We trained 3 types of
machine learning models (random forest, logistic regression, and support
vector machine) with bacterial community data from 5 days post-challenge
to predict whether the mice were still colonized with \emph{C.
difficile} 10 days post-challenge. We chose 5 days post-challenge
because that was the earliest time point where we would see a treatment
effect in the mice that were given 1-day PEG treatment three days after
challenge. The random forest model had the highest performance (AUROC =
0.90, Data Set S1, Sheet X), so we next performed permutation importance
to examine the bacteria that were the top contributors to the random
forest model predicting prolonged \emph{C. difficile }colonization. We
selected the top 10 bacteria contributing to our models performance
(Fig. 7A) and examined their relative abundance at 5 days
post-challenge, the time point used to predict \emph{C. difficile}
colonization status on day 10 (Fig. 7B). Next, we focused on the 5
genera that had a greater than 1 \% relative abundance in either the
cleared or colonized mice and examined how the bacteria changed over
time. We found \emph{Enterobacteriaceae} and \emph{Bacteroides} tended
to have a higher relative abundance, the relative abundance of
\emph{Akkermansia} was initially decreased and then increased, and
\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} had a lower
relative abundance in the mice with prolonged colonization compared to
the mice that cleared \emph{C. difficile} (Fig. 7C). Together these
results suggest a combination of low and high abundance bacterial genera
influence the prolonged colonization observed in 5-day PEG and post-CDI
PEG treated mice.

Previous work examining the impact of PEG on the murine microbiota found
that PEG treatment resulted in the permanent loss of \emph{Muribaculum
intestinale} (16)., which is classified as \emph{Porphyromonadaceae} by
the Ribosomal Database Project (RDP) database (31).
\emph{Porphyromonadaceae} was a top feature in the random forest model
predicting prolonged \emph{C. difficile} colonization. We identified 4
OTUs that had 92-96\% identity to \emph{Muribaculum intestinale} and
examined their abundance in mice that either cleared or were still
colonized with \emph{C. difficile} at 10 days post-challenge. While all
of the OTUs, were decreased by PEG and clindamycin treatment, there was
some recovery in the mice that cleared (Fig. S4A). We also examined
other \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs since
these were the 2 genera that were important to our classification model
and contained multiple OTUs that were different at 5 days post-challenge
between mice that either cleared or remained colonized with \emph{C.
difficile} by 10 days post-challenge (Data Set S1, sheet X). While
individual \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs
tended to be more abundant in the mice that clear \emph{C. difficile}
relative to the mice that exhibit prolonged colonization (Fig. S4B-C),
there is no single OTU that fits the pattern we observed at the genus
level (Fig. 7C), suggesting multiple \emph{Porphyromonadaceae} and
\emph{Lachnospiraceae} OTUs influenced \emph{C. difficile} clearance.
Overall, our results suggest that specific bacterial community
differences explain the prolonged \emph{C. difficile} colonization we
observed in 5-day PEG and post-CDI 1-day PEG treated mice.

@@

\begin{itemize}
\item
  Summary of major findings (Fig. 8A)
\item
  Discussion of prolonged persistence. \emph{C. difficile} sequences
  detected in tissue samples. Association with mucin-degrading bacteria
  suggested by recent papers.
\item
  Discuss why we might not have observed more severe histology in PEG
  mice relative to clindamycin-treated mice

  \begin{itemize}
  \tightlist
  \item
    Antibiotics may also impact mucus layer
  \item
    Strain of bacteria used
  \end{itemize}
\item
  Protective bacteria missing in PEG-treated mice
\item
  Discuss what these findings might mean for human patients (Fig. 8B)

  \begin{itemize}
  \tightlist
  \item
    What's known regarding laxatives and susceptibility to CDIs
  \item
    Clinical trial of PEG, results never posted (32)
  \item
    Relevance to human FMTs? Unclear what the best administration route
    is because there have been no studies designed to evaluate the best
    administration route for FMTs.
  \end{itemize}
\end{itemize}

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

>>>>

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because of their impact on the microbiota. However,
non-antibiotic medications such as the ubiquitous osmotic laxative
polyethylene glycol (PEG) 3350, also alter the microbiota, but whether
PEG impacts CDI susceptibility and clearance is unclear. To examine how
PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day
doses of 15\% PEG in the drinking water and then challenged the mice
with \emph{C. difficile} 630. We used clindamycin-treated mice as a
control because they consistently clear \emph{C. difficile} within 10
days post-challenge (dpc). PEG treatment alone was sufficient to render
mice susceptible and 5-day PEG-treated mice remain colonized for up to
30 dpc. In contrast, 1-day PEG treated mice were transiently colonized,
clearing \emph{C. difficile} within 7 dpc. To examine how PEG treatment
impacts clearance, we administered a 1-day PEG treatment to
clindamycin-treated, \emph{C. difficile}-challenged mice. Administering
PEG to mice after \emph{C. difficile} challenge prolonged colonization
up to 30 dpc. When we trained a random forest model with community data
from 5 dpc, we were able to predict which mice would exhibit prolonged
colonization (AUROC = 0.90). Five of the top ten bacterial features
important for predicting prolonged colonization had high relative
abundances in the community. Examining the dynamics of these bacterial
during the post-challenge period revealed patterns in the relative
abundances of \emph{Bacteroides}, \emph{Enterobacteriaceae},
\emph{Porphyromonadaceae}, \emph{Lachnospiraceae}, and
\emph{Akkermansia} that were associated with prolonged \emph{C.
difficile} colonization in PEG-treated mice.

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

Diarrheal samples induced by medications such as laxatives are typically
rejected for \emph{Clostridiodes difficile} testing. However, there are
some microbiota similarities between diarrheal and \emph{C. difficile}
infection (CDI) communities such as lower diversity compared to healthy
controls, which lead us to hypothesize that diarrhea may be an indicator
of \emph{C. difficile} risk. We explored how osmotic laxatives disrupt
the microbiota's colonization resistance to \emph{C. difficile} by
administering a laxative to mice either before or after \emph{C.
difficile} challenge Our findings suggest the osmotic laxative disrupts
colonization resistance to \emph{C. difficile}, as well as clearance in
mice already colonized with \emph{C. difficile}. Considering that most
hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact
microbiota colonization resistance in humans.

@@

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because they disrupt microbiota colonization
resistance (1). However, antibiotics are not the only types of
medications that disrupt the microbiota (2--4). Although, other
medications (proton pump inhibitors, osmotic laxatives, antimotility
agents, and opioids) have been implicated as risk or protective factors
for CDIs through epidemiological studies, whether the association is due
to their impact on the microbiota is still unclear (5--9).

@@

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although, two groups have now shown PEG treatment
alone renders mice susceptible to \emph{C. difficile} (15, 17), these
studies have raised additional questions regarding the dynamics and
severity of infection as well as the role of laxative treatment in
\emph{C. difficile} clearance. Addressing these questions will better
inform how we think about laxatives in the context of CDIs. Here, we
characterized how long PEG-treated mice remain susceptible, whether PEG
treatment results in more sustained \emph{C. difficile} colonization and
severe CDI than mice treated with clindamycin, and if PEG treatment
after challenge can promote \emph{C. difficile} clearance.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders the mice susceptible to \emph{C. difficile} colonization, with
mice typically clearing \emph{C. difficile} within 10 days
post-challenge (9, 26). All three groups of PEG-treated mice were
administered a 15\% PEG solution in the drinking water for 5-days: 1.
received no additional treatment, 2. was also treated with clindamycin,
and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A).
PEG treatment resulted in weight loss in all 3 groups of PEG-treated
mice relative to their baseline weights, with the greatest change in
weight observed on the fifth day of PEG treatment. The mice recovered
most of the lost weight by five days after treatment (Fig. 1B). After
either the PEG, clindamycin, or PEG and clindamycin treatment all mice
were challenged with 10\textsuperscript{5} \emph{C. difficile} 630
spores (Fig. 1A). All treatments rendered mice susceptible to \emph{C.
difficile} colonization. However, PEG-treated mice remained colonized
with \emph{C. difficile} at a high level through thirty days
post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice
cleared \emph{C. difficile} within ten days post-challenge (Fig. 1C).
Therefore, PEG treatment led to sustained colonization in contrast to
clindamycin mice that naturally cleared \emph{C. difficile} within ten
days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to
\emph{C. difficile} colonization after a 10-day recovery period,
although \emph{C. difficile} was not detectable in most of the group in
the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
dead on the 6th day post-challenge, presumably due to \emph{C.
difficile} as the bacteria became detectable in stool samples from that
mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days
post-challenge onward, the density of \emph{C. difficile} stabilized in
the 10-day recovery group and remained high through 20-30 days
post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was
sufficient to render mice susceptible to prolonged \emph{C. difficile}
colonization and PEG-treated mice remained susceptible through ten days
post-treatment.

\textbf{5-day laxative treatment differentially disrupted the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation plate
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to coprophagy (27). We tried to
minimize the impact of cage effects on our experiment by breaking up
littermates when assigning mice to treatment groups. Importantly,
although we conducted a total of 5 separate experiments, the experiment
number and its interaction with treatment group did no significantly
explain the observed variation in fecal communities (Data Set S1, Sheet
X). None of the treatment groups recovered to their baseline community
structure either 10 or 30 days post-challenge suggesting other community
features besides recovery to baseline were responsible for the prolonged
\emph{C. difficile} colonization in PEG-treated mice (Fig. 2B).

@@

Interestingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine if there were genera that
changed during the post-challenge period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-challenge to
detectable in the stool samples with the density stabilizing around 8
days post-challenge (Fig. S1A). We found no bacteria with relative
abundances that were significantly different over the two time points
after multiple hypothesis correction (Data Set S1, Sheet X). However,
there was also wide variation between individual mice regarding when
\emph{C. difficile} became detectable (Fig. S1A) as well as the relative
abundances of bacterial genera in the communities (Fig. S1B). For
example, two mice had a high relative abundance of
\emph{Enterobacteriaceae} throughout the post-challenge period and this
corresponded to mouse 10, which died on the sixth day post-challenge and
mouse 11, where \emph{C. difficile} was present at a high density from
the 4th day post-challenge onward (Fig. S1B). While we did not identify
a clear signal to explain the delayed appearance of \emph{C. difficile}
in the 5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment did not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced snips of tissue collected
from the cecum, proximal colon, and distal colon. Similar to what was
observed with the stool samples, alpha diversity was lower in the
PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
diversity continued to increase over time based on the tissues from
PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A,
Data Set S1, Sheet X). Group, time point, and their interactions with
other variables (cage, experiment number, and sample type) explained the
majority of the variation observed in mucosal communities (PERMANOVA
combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
3B, Data Set S1, Sheet X). We saw the greatest difference in the
relative abundance of the mucosal microbiota between treatment groups
(clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
post-challenge with 10 genera that were significantly different
(\emph{P} \textless{} 0.05) in all three of the tissue types we
collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set
S1, Sheet X). Interestingly, \emph{Peptostreptococcaceae} (the family
with a sequence that matches \emph{C. difficile}) was one of the genera
that had a significant difference in relative abundance between
treatment groups at 6 days post-challenge. This population was primarily
only present in the 5-day PEG treatment group of mice and decreased in
the proximal and distal colon tissues over time (Fig. S2B). By 30 days
post-challenge, only the relative abundances of \emph{Bacteroides},
\emph{Clostridiales}, \emph{Firmicutes}, and \emph{Ruminococcaceae} were
different between treatment groups and only in the cecum tissues (Fig.
3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a
significant impact on the mucosal microbiota and we detected \emph{C.
difficile} sequences in the cecum, proximal colon, and distal colon
tissue communities.

Because there were differences in the mucosal microbiota including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
H\&E stained histopathology (28). However, we found there was no
difference in cecum and colon scores between clindamycin and PEG-treated
mice that were challenged with \emph{C. difficile} at 4 days
post-challenge (Fig. 3D), the time point typically examined in \emph{C.
difficile} 630 challenged mice (29). We also looked at 6 days
post-challenge because that was when there was a large difference in
\emph{C. difficile} density between PEG- and clindamycin-treated mice
(Fig. 1C). Although, there was a slight difference in the colon between
PEG and clindamycin-treated mice, there was not a signifant difference
in the cecum and the overall score was relatively low (1.5-2.5 out of
12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect
on the mucosal microbiota, the impact of \emph{C. difficile} 630
challenge on the cecum and colon was similar between PEG and clindamycin
treated mice.

@@

\textbf{Post-challenge laxative treatment disrupted clearance in
clindamycin-treated mice regardless of whether an FMT was also
administered.} Since a 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. To examine the impact of PEG treatment on
\emph{C. difficile} clearance, we treated 4 groups of mice with
clindamycin and then challenged all mice with \emph{C. difficile} before
administering the following treatments: no additional treatment, 1-day
PEG immediately after challenge, and 1-day PEG treatment 3 days after
challenge followed by either administration of an FMT or PBS solution by
oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice
that received PEG exhibited prolonged \emph{C. difficile} colonization
(Fig. 5B).

@@

\textbf{Five-day post-challenge community data can predict which mice
that will have prolonged \emph{C. difficile} colonization.} After
identifying bacteria associated with the 5-day, 1-day and post-challenge
1-day PEG treatments, we decided to examine the bacteria that influenced
prolonged \emph{C. difficile} colonization. We trained 3 types of
machine learning models (random forest, logistic regression, and support
vector machine) with bacterial community data from 5 days post-challenge
to predict whether the mice were still colonized with \emph{C.
difficile} 10 days post-challenge. We chose 5 days post-challenge
because that was the earliest time point where we would see a treatment
effect in the post-challenge 1-day PEG experiments. The random forest
model had the highest performance (median AUROC = 0.90, Data Set S1,
Sheet X), so we next performed permutation importance to examine the
bacteria that were the top contributors to the random forest model
predicting prolonged \emph{C. difficile }colonization. We selected the
top 10 bacteria contributing to our model's performance (Fig. 7A) and
examined their relative abundance at 5 days post-challenge, the time
point used to predict \emph{C. difficile} colonization status on day 10
(Fig. 7B). Next, we focused on the 5 genera that had a greater than 1 \%
relative abundance in either the cleared or colonized mice and examined
how the bacteria changed over time. We found \emph{Enterobacteriaceae}
and \emph{Bacteroides} tended to have a higher relative abundance, the
relative abundance of \emph{Akkermansia} was initially decreased and
then increased, and \emph{Porphyromonadaceae} and \emph{Lachnospiraceae}
had a lower relative abundance in the mice with prolonged colonization
compared to the mice that cleared \emph{C. difficile} (Fig. 7C).
Together these results suggest a combination of low and high abundance
bacterial genera influence the prolonged colonization observed in 5-day
PEG and post-challenge 1-day PEG treated mice.

@@

While the disruptive effect of antibiotics on \emph{C. difficile}
colonization resistance is well established, the extent to which other
drugs such as laxatives disrupt colonization resistance was unclear. By
following osmotic laxative treated mice over time, we found 5-day PEG
treatment before challenge resulted in prolonged \emph{C. difficile}
colonization, while a 1-day PEG treatment resulted in transient
colonization. The differences in \emph{C. difficile} colonization
dynamics between the 5- and 1-day PEG treated mice were associated with
differences in how much the treatments disrupted the microbiota.
Additionally, the intestinal communities of 5-day PEG treated mice had
not regained colonization resistance after a 10-day recovery period.
Although in contrast to the other 5-day PEG treatment groups, \emph{C.
difficile} was not immediately detectable in the stools of most of the
mice in the 10-day recovery group. We also examined the impact of PEG
treatment after \emph{C. difficile} challenge and in opposition to the
hypothesis suggested by the literature, found that PEG treatment
prolonged colonization relative to mice that only recieved clindamycin
treatment. We identified patterns in the relative abundances of
\emph{Bacteroides}, \emph{Enterobacteriaceae}, \emph{Akkermansia},
\emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
associated with prolonged \emph{C. difficile} colonization (Fig. 8).
Overall, our results demonstrated that osmotic laxative treatment alone
rendered mice susceptible to \emph{C. difficile} colonization and the
duration of colonization depended on the length of PEG treatment and
whether treatment was administered before or after challenge.

In addition to altering composition, laxative treatment may alter
microbiota-produced metabolites. A previous study demonstrated that a
5-day treatment of 10\% PEG depleted acetate and butyrate and increased
succinate compared to untreated mice (15). While we did not perform
metabolomic analysis, we did see bacteria known to produce benefical
metabolites were depleted in mice that cleared \emph{C. difficile}
compared to mice with prolonged colonization. For example,
\emph{Oscillibacter valericigenes} can produce the SCFA valerate (31),
and separate studies demonstrated valerate is depleted after clindamycin
treatment and inhibited \emph{C. difficile} growth based on \emph{in
vitro} data and an experiment with C57BL/6 mice (32, 33). Similarly,
\emph{Acetatifactor} can produce acetate and butyrate (34), SCFAs that
are decreased in mice with prolonged \emph{C. difficile} infection after
antibiotic treatment (35). Thus protective bacteria and their
metabolites could be depleted by osmotive laxative treatment depending
on the timing and duration of treatment.
